---
title: "Model Comparison & Final Report"
subtitle: "Comprehensive Analysis of Regression Portfolio"
author: "Osman Omer Mustafa"
date: "today"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
    toc-location: left
execute:
  warning: false
  message: false
---

## Executive Summary

This report compares all regression models developed across Projects 1-7.

**Datasets:**

- Insurance (n=1338): Predicting medical charges
- Heart Disease (n=1025): Predicting disease presence

**Key Findings:**

- Linear: Multiple regression best (R² = 0.75)
- Logistic: Full model best (AUC = 0.95)
- Regularization: Similar performance (AUC ≈ 0.92)

## Load Libraries

```{r setup}
library(tidyverse)
library(broom)
library(pROC)
library(caret)
library(glmnet)
```

## Part 1: Linear Regression Models

### Data Preparation

```{r}
insurance <- read_csv("insurance.csv")

insurance <- insurance %>%
  mutate(
    sex = as.factor(sex),
    smoker = as.factor(smoker),
    region = as.factor(region)
  )
```

### Model 1: Simple Linear

Predicting charges from age only.

```{r}
model_simple <- lm(charges ~ age, data = insurance)
glance_simple <- glance(model_simple)
```

### Model 2: Multiple Linear

Including age, BMI, smoker status, and region.

```{r}
model_multiple <- lm(charges ~ age + bmi + smoker + region, data = insurance)
glance_multiple <- glance(model_multiple)
```

### Model 3: Polynomial

Adding quadratic terms for age and BMI.

```{r}
model_poly <- lm(charges ~ age + I(age^2) + bmi + I(bmi^2), data = insurance)
glance_poly <- glance(model_poly)
```

### Linear Models Comparison

```{r}
linear_comparison <- data.frame(
  Model = c("Simple", "Multiple", "Polynomial"),
  R_squared = c(glance_simple$r.squared, glance_multiple$r.squared, glance_poly$r.squared),
  Adj_R_squared = c(glance_simple$adj.r.squared, glance_multiple$adj.r.squared, glance_poly$adj.r.squared),
  AIC = c(AIC(model_simple), AIC(model_multiple), AIC(model_poly)),
  BIC = c(BIC(model_simple), BIC(model_multiple), BIC(model_poly))
)

linear_comparison %>% knitr::kable(caption = "Linear Models Performance", digits = 3)
```

```{r echo=FALSE}
best_r2 <- max(linear_comparison$R_squared)
best_model_linear <- linear_comparison$Model[which.max(linear_comparison$R_squared)]
```

**Winner:** `r best_model_linear` (R² = `r round(best_r2, 3)`)

### Visualization

```{r}
ggplot(linear_comparison, aes(x = Model, y = R_squared, fill = Model)) +
  geom_bar(stat = "identity") +
  labs(title = "Linear Models: R² Comparison", y = "R²") +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```

## Part 2: Logistic Regression Models

### Data Preparation

```{r}
heart <- read_csv("heart.csv")

heart <- heart %>%
  mutate(
    target = as.factor(target),
    sex = as.factor(sex),
    cp = as.factor(cp),
    fbs = as.factor(fbs),
    restecg = as.factor(restecg),
    exang = as.factor(exang),
    slope = as.factor(slope),
    ca = as.factor(ca),
    thal = as.factor(thal)
  )
```

### Model 1: Baseline (Intercept-only)

```{r}
model_baseline <- glm(target ~ 1, data = heart, family = binomial)
pred_baseline <- predict(model_baseline, type = "response")
class_baseline <- ifelse(pred_baseline > 0.5, "1", "0")
auc_baseline <- auc(roc(heart$target, pred_baseline))
conf_baseline <- confusionMatrix(as.factor(class_baseline), heart$target, positive = "1")
```

### Model 2: Clinical Model

```{r}
model_clinical <- glm(target ~ age + sex + cp + trestbps + chol + fbs + thalach,
                      data = heart, family = binomial)
pred_clinical <- predict(model_clinical, type = "response")
class_clinical <- ifelse(pred_clinical > 0.5, "1", "0")
auc_clinical <- auc(roc(heart$target, pred_clinical))
conf_clinical <- confusionMatrix(as.factor(class_clinical), heart$target, positive = "1")
```

### Model 3: Full Model

```{r}
model_full <- glm(target ~ ., data = heart, family = binomial)
pred_full <- predict(model_full, type = "response")
class_full <- ifelse(pred_full > 0.5, "1", "0")
auc_full <- auc(roc(heart$target, pred_full))
conf_full <- confusionMatrix(as.factor(class_full), heart$target, positive = "1")
```

### Logistic Models Comparison

```{r}
logistic_comparison <- data.frame(
  Model = c("Baseline", "Clinical", "Full"),
  AIC = c(AIC(model_baseline), AIC(model_clinical), AIC(model_full)),
  AUC = c(auc_baseline, auc_clinical, auc_full),
  Accuracy = c(
    conf_baseline$overall["Accuracy"],
    conf_clinical$overall["Accuracy"],
    conf_full$overall["Accuracy"]
  ),
  Sensitivity = c(
    conf_baseline$byClass["Sensitivity"],
    conf_clinical$byClass["Sensitivity"],
    conf_full$byClass["Sensitivity"]
  ),
  Specificity = c(
    conf_baseline$byClass["Specificity"],
    conf_clinical$byClass["Specificity"],
    conf_full$byClass["Specificity"]
  )
)

logistic_comparison %>% knitr::kable(caption = "Logistic Models Performance", digits = 3)
```

```{r echo=FALSE}
best_auc_log <- max(logistic_comparison$AUC)
best_model_log <- logistic_comparison$Model[which.max(logistic_comparison$AUC)]
```

**Winner:** `r best_model_log` (AUC = `r round(best_auc_log, 3)`)

### ROC Curves

```{r}
plot(roc(heart$target, pred_baseline), col = "gray", lwd = 2, main = "Logistic Models: ROC Comparison")
lines(roc(heart$target, pred_clinical), col = "blue", lwd = 2)
lines(roc(heart$target, pred_full), col = "red", lwd = 2)
legend("bottomright",
       legend = c("Baseline", "Clinical", "Full"),
       col = c("gray", "blue", "red"), lwd = 2)
```

## Part 3: Regularized Regression

### Data Preparation

```{r}
x <- model.matrix(target ~ ., data = heart)[, -1]
y <- as.numeric(as.character(heart$target))
```

### Ridge Regression

```{r}
set.seed(123)
cv_ridge <- cv.glmnet(x, y, family = "binomial", alpha = 0)
model_ridge <- glmnet(x, y, family = "binomial", alpha = 0, lambda = cv_ridge$lambda.min)
pred_ridge <- as.vector(predict(model_ridge, newx = x, type = "response"))
auc_ridge <- auc(roc(y, pred_ridge))
```

### Lasso Regression

```{r}
set.seed(123)
cv_lasso <- cv.glmnet(x, y, family = "binomial", alpha = 1)
model_lasso <- glmnet(x, y, family = "binomial", alpha = 1, lambda = cv_lasso$lambda.min)
pred_lasso <- as.vector(predict(model_lasso, newx = x, type = "response"))
auc_lasso <- auc(roc(y, pred_lasso))
```

### Elastic Net

```{r}
set.seed(123)
cv_elastic <- cv.glmnet(x, y, family = "binomial", alpha = 0.5)
model_elastic <- glmnet(x, y, family = "binomial", alpha = 0.5, lambda = cv_elastic$lambda.min)
pred_elastic <- as.vector(predict(model_elastic, newx = x, type = "response"))
auc_elastic <- auc(roc(y, pred_elastic))
```

### Regularized Models Comparison

```{r}
regularized_comparison <- data.frame(
  Model = c("Ridge", "Lasso", "Elastic Net"),
  AUC = c(auc_ridge, auc_lasso, auc_elastic),
  N_Predictors = c(
    sum(coef(model_ridge) != 0) - 1,
    sum(coef(model_lasso) != 0) - 1,
    sum(coef(model_elastic) != 0) - 1
  )
)

regularized_comparison %>% knitr::kable(caption = "Regularized Models Performance", digits = 4)
```

```{r echo=FALSE}
best_auc_reg <- max(regularized_comparison$AUC)
best_model_reg <- regularized_comparison$Model[which.max(regularized_comparison$AUC)]
```

**Winner:** `r best_model_reg` (AUC = `r round(best_auc_reg, 3)`)

## Overall Comparison

### Best Models Summary

```{r}
summary_table <- data.frame(
  Task = c("Linear Regression", "Logistic Regression", "Regularized Regression"),
  Best_Model = c(best_model_linear, best_model_log, best_model_reg),
  Metric = c(
    paste0("R² = ", round(best_r2, 3)),
    paste0("AUC = ", round(best_auc_log, 3)),
    paste0("AUC = ", round(best_auc_reg, 3))
  )
)

summary_table %>% knitr::kable(caption = "Best Models Summary")
```

## Key Insights

### Linear Regression (Insurance Data)

**Multiple regression performed best** (R² = `r round(best_r2, 3)`), explaining 
`r round(best_r2 * 100, 1)`% of variance in medical charges.

**Key predictors:** Smoking status had the strongest effect (~$23,000 increase).

**Polynomial terms** did not substantially improve performance, suggesting 
linear relationships are adequate.

### Logistic Regression (Heart Disease)

**Full model achieved excellent performance** (AUC = `r round(best_auc_log, 3)`), 
with `r round(best_auc_log * 100, 1)`% discriminative ability.

**Sensitivity:** `r round(conf_full$byClass["Sensitivity"], 3)` 
(detects `r round(conf_full$byClass["Sensitivity"] * 100, 1)`% of disease cases)

**Improvement over baseline:** 
AUC increased from `r round(auc_baseline, 3)` to `r round(best_auc_log, 3)`.

### Regularization

All three methods performed similarly (AUC ≈ 0.92), indicating:

- Dataset is well-behaved
- No severe multicollinearity
- Moderate number of predictors

**Regularization most useful** when dealing with high-dimensional data 
(hundreds/thousands of predictors).

## Recommendations

### For Medical Charge Prediction

**Use multiple linear regression** with key predictors (age, BMI, smoker, region).

Provides good accuracy (R² = 0.75) with interpretable coefficients.

### For Heart Disease Diagnosis

**Use full logistic regression model** including all available diagnostic tests.

Achieves 95% AUC with 92% sensitivity (few missed cases).

**For screening:** Clinical model (AUC = 0.88) using basic measurements may suffice.

### Methodology Best Practices

**Model comparison should include:**

- Multiple metrics (R², AIC, BIC for linear; AUC, sensitivity, specificity for logistic)
- Visual comparisons (ROC curves, coefficient plots)
- Consideration of complexity vs performance trade-off

**Always validate on independent data** before clinical application.

## Limitations

**No train-test split:** All models evaluated on training data, potentially 
overestimating performance.

**Single datasets:** Results specific to insurance and heart disease data.

**No external validation:** Required for clinical deployment.

**Educational purpose:** These analyses demonstrate methodology, not clinical tools.

## Conclusion

This portfolio successfully demonstrates:

✅ Linear regression techniques (simple, multiple, polynomial)

✅ Logistic regression for binary classification

✅ Regularization methods (Ridge, Lasso, Elastic Net)

✅ Model comparison and evaluation

✅ Professional reporting and interpretation

**The progression from simple to complex models**, combined with proper 
evaluation and comparison, provides a solid foundation for applied regression 
analysis in medical and research contexts.

**Future directions** include cross-validation, feature engineering, 
and exploration of machine learning methods (random forests, neural networks).
